---
title: "Reproducible Research: Peer Assessment 1"
author: "ryaxley"
date: "February 28, 2016"
output: html_document
keep_md: true
---


## Loading and preprocessing the data
```{r}
setwd("~/Reproducible.Research/code/Project.1/RepData_PeerAssessment1/")
df <- read.csv(unz("activity.zip", "activity.csv"), header=T, sep=",")

# Convert date variable to date format
df$date <- as.Date(df$date, format="%Y-%m-%d")

# Convert interval variable to factor
df$interval <- as.factor(df$interval)
```


## What is mean total number of steps taken per day?
```{r}
# Aggregate steps by day
daily <- aggregate(steps ~ date, df, sum)

# Total steps by day (sorted)
sort(daily$steps, decreasing = TRUE)

# Histogram of daily steps
hist(daily$steps, breaks = 5)

# Mean steps per day
mean(daily$steps)

# Median steps per day
median(daily$steps)
```


## What is the average daily activity pattern?
```{r}
# Exploratory plot of avg steps across days
plot(daily$date, daily$steps, type="l", col='red', 
     xlab="Date", ylab="Total steps taken each day")
abline(h=mean(daily$steps), lwd=1, lty=2 )

# Mean steps by 5 minute intervals
interval <- aggregate(steps ~ interval, df, mean)
plot(interval$steps, type="l", 
     xlab="5 minute intervals during the day", ylab="Average steps")

# Maximum number of steps per 5-m interval across days
max(interval$steps)
abline(v=which.max(interval$steps), col="red", lty=2)
text(200,170, paste("Maximum average steps =", round(max(interval$steps))))
```

## Imputing missing values
With the MICE package we imputed missing values. The overall pattern remains very similar to the original data. 
```{r}
# Number of missing values
length(subset(df, is.na(steps))$steps)

# Impute library: Multivariate Imputation by Chained Equations  
library(mice)
df2 <- df
# Date variable needs to be converted for MICE to run
df2$date <- as.character(df2$date)
# Impute data
temp <- mice(df2, m=5, maxit=10, meth='pmm', seed=500)
df2 <- complete(temp, include=TRUE)

# Aggregate steps by day
daily2 <- aggregate(steps ~ date, df2, sum)

# Histogram of daily steps
hist(daily2$steps, breaks = 5)

# Mean steps per day
mean(daily2$steps)

# Median steps per day
median(daily2$steps)

# Mean steps by 5 minute intervals
interval2 <- aggregate(steps ~ interval, df2, mean)
plot(interval2$steps, type="l", 
     xlab="5 minute intervals during the day", ylab="Average steps")

# Maximum number of steps per 5-m interval across days
max(interval2$steps)
abline(v=which.max(interval2$steps), col="red", lty=2)
text(200,170, paste("Maximum steps =", round(max(interval2$steps))))
```


## Are there differences in activity patterns between weekdays and weekends?
Segregating the data by day of the week we can see a larger number of steps distributed throughout the entire day compared to the weekdays.
```{r}

# Create a new factor variable for day of the week
df2$day <- NA
df2$date <- as.Date(df2$date)
df2$day <- weekdays(df2$date)
df2[df2$day == 'Monday', ]$day <- 'weekday'
df2[df2$day == 'Tuesday', ]$day <- 'weekday'
df2[df2$day == 'Wednesday', ]$day <- 'weekday'
df2[df2$day == 'Thursday', ]$day <- 'weekday'
df2[df2$day == 'Friday', ]$day <- 'weekday'
df2[df2$day == 'Saturday', ]$day <- 'weekend'
df2[df2$day == 'Sunday', ]$day <- 'weekend'

df2 <- aggregate(steps ~ day+interval, df2, mean)

library(lattice)

xyplot(steps~interval|factor(day), data=df2,
       type='l', layout=c(1,2),
       xlab='Interval', ylab='Number of Steps')


```
